---
title: "Gov 50 Final Project"
author: "Anita Carew"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project proposal

 My research question for this project is: Does a country's GDP increase its carbon emissions? The widely held belief by political and climate scientists around the world has long been that the higher a country's GDP, the more their carbon emissions increase. This goes along with the idea that if a country is more economicaly developed, ergo more technologically advanced with a more elaborate quality of lifestyle for its citizens, their waste and consumption will lead to a higher carbon emission within said country. There is only a little evidence such as the 2016 report by the International Energy Agency that claims some countries do exist that have been able to reduce Green House Gas (GHG) emissions while growing GDP. Nevertheless, this idea assumes that a country's technology and economy is at a stable enough level where they can put money into climate conscious policy. Most carbon emissions-GDP studies discover that countries usually spend the most carbon emissions when they are attempting to increase economic growth considerably. Thus, I expect that GDP and carbon emissions per capita have a positive relationship. 

  My data is derived from the World Bank. My sample is made up of all the countries the World Bank measures, including some territories (total: 217). The independent variable is GDP per capita, PPP (current international $). This is measured by using the gross domestic product (GDP) expressed in current international dollars converted by purchasing power parity (PPP) conversion factor. The dependent variable is carbon emissions which we will assume rise as a result of higher GDP. This dependent variable is measures by carbon dioxide emissions (metric tons per capita), which is seen through emissions which are stemming from the burning of fossil fuels and the manufacture of cement. They include carbon dioxide produced during consumption of solid, liquid, and gas fuels and gas flaring.
  
  My outcome variable is the correlation relationship between this GDP and CO2 emissions. If the correlational variable is large and positive, then that means my hypothesis has been proven correct. If there seems to be no relationship between the variables (correlational coefficient is close to zero), that suggests economic growth has no effect on the emissions caused by higher production and consumption within a country. My hypothesis would also be wrong if this variable is large and negative, which implies GDP has a negative effect on emissions, and possibly that growth in GDP allows for technological growth that accounts for using more clean energy on a wider scale. I might also look into creating separate visualizations of the dataset in case analysis shows that more developed countries or that countries from a certain region have a much slower or faster slope of C02 and carbon emissions. 


```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(readr)
final_wb_data <- read_csv("final_wb_data.csv")
final_wb_data



final_wb_data$`1990 [YR1990]`<-as.numeric(as.character(final_wb_data$`1990 [YR1990]`))
final_wb_data$`2022 [YR2022]`<-as.numeric(as.character(final_wb_data$`2022 [YR2022]`))
final_wb_data$`1991 [YR1991]`<-as.numeric(as.character(final_wb_data$`1991 [YR1991]`))
final_wb_data$`1992 [YR1992]`<-as.numeric(as.character(final_wb_data$`1992 [YR1992]`))
final_wb_data$`1993 [YR1993]`<-as.numeric(as.character(final_wb_data$`1993 [YR1993]`))
final_wb_data$`1994 [YR1994]`<-as.numeric(as.character(final_wb_data$`1994 [YR1994]`))
final_wb_data$`1995 [YR1995]`<-as.numeric(as.character(final_wb_data$`1995 [YR1995]`))
final_wb_data$`1996 [YR1996]`<-as.numeric(as.character(final_wb_data$`1996 [YR1996]`))
final_wb_data$`1997 [YR1997]`<-as.numeric(as.character(final_wb_data$`1997 [YR1997]`))
final_wb_data$`1998 [YR1998]`<-as.numeric(as.character(final_wb_data$`1998 [YR1998]`))
final_wb_data$`1999 [YR1999]`<-as.numeric(as.character(final_wb_data$`1999 [YR1999]`))
final_wb_data$`2000 [YR2000]`<-as.numeric(as.character(final_wb_data$`2000 [YR2000]`))
final_wb_data$`2001 [YR2001]`<-as.numeric(as.character(final_wb_data$`2001 [YR2001]`))
final_wb_data$`2002 [YR2002]`<-as.numeric(as.character(final_wb_data$`2002 [YR2002]`))
final_wb_data$`2003 [YR2003]`<-as.numeric(as.character(final_wb_data$`2003 [YR2003]`))
final_wb_data$`2004 [YR2004]`<-as.numeric(as.character(final_wb_data$`2004 [YR2004]`))
final_wb_data$`2005 [YR2005]`<-as.numeric(as.character(final_wb_data$`2005 [YR2005]`))
final_wb_data$`2006 [YR2006]`<-as.numeric(as.character(final_wb_data$`2006 [YR2006]`))
final_wb_data$`2007 [YR2007]`<-as.numeric(as.character(final_wb_data$`2007 [YR2007]`))
final_wb_data$`2008 [YR2008]`<-as.numeric(as.character(final_wb_data$`2008 [YR2008]`))
final_wb_data$`2009 [YR2009]`<-as.numeric(as.character(final_wb_data$`2009 [YR2009]`))
final_wb_data$`2010 [YR2010]`<-as.numeric(as.character(final_wb_data$`2010 [YR2010]`))
final_wb_data$`2011 [YR2011]`<-as.numeric(as.character(final_wb_data$`2011 [YR2011]`))
final_wb_data$`2012 [YR2012]`<-as.numeric(as.character(final_wb_data$`2012 [YR2012]`))
final_wb_data$`2013 [YR2013]`<-as.numeric(as.character(final_wb_data$`2013 [YR2013]`))
final_wb_data$`2014 [YR2014]`<-as.numeric(as.character(final_wb_data$`2014 [YR2014]`))
final_wb_data$`2015 [YR2015]`<-as.numeric(as.character(final_wb_data$`2015 [YR2015]`))
final_wb_data$`2016 [YR2016]`<-as.numeric(as.character(final_wb_data$`2016 [YR2016]`))
final_wb_data$`2017 [YR2017]`<-as.numeric(as.character(final_wb_data$`2017 [YR2017]`))
final_wb_data$`2018 [YR2018]`<-as.numeric(as.character(final_wb_data$`2018 [YR2018]`))
final_wb_data$`2019 [YR2019]`<-as.numeric(as.character(final_wb_data$`2019 [YR2019]`))
final_wb_data$`2020 [YR2020]`<-as.numeric(as.character(final_wb_data$`2020 [YR2020]`))
final_wb_data$`2021 [YR2021]`<-as.numeric(as.character(final_wb_data$`2021 [YR2021]`))



ranks<- final_wb_data|>
  filter(`Series Name`== "GDP per capita, PPP (current international $)")|>
  group_by(`Country Name`)|>
  mutate(totalgdp= sum(`1990 [YR1990]`, `1991 [YR1991]`, `1992 [YR1992]`, `1993 [YR1993]`, `1994 [YR1994]`, `1995 [YR1995]`, `1996 [YR1996]`, `1997 [YR1997]`, `1998 [YR1998]`, `1999 [YR1999]`, `2000 [YR2000]`, `2001 [YR2001]`, `2002 [YR2002]`, `2003 [YR2003]`, `2004 [YR2004]`, `2005 [YR2005]`, `2006 [YR2006]`, `2007 [YR2007]`, `2008 [YR2008]`, `2009 [YR2009]`, `2010 [YR2010]`, `2011 [YR2011]`, `2012 [YR2012]`, `2013 [YR2013]`, `2014 [YR2014]`, `2015 [YR2015]`, `2016 [YR2016]`, `2017 [YR2017]`, `2018 [YR2018]`, `2019 [YR2019]`, `2020 [YR2020]`,  `2021 [YR2021]`, `2022 [YR2022]`, na.rm=TRUE))

ranks




```


##Rank top 20 of data 'total gdp' column
```{r}



ranks <- final_wb_data[order(desc(final_wb_data$`2019 [YR2019]`)),]
ranks <- ranks[1:20,]


ranks

```



##Pivot longer
```{r}
copy<- ranks
copy<- copy|>
  pivot_longer(cols = `1990 [YR1990]` : `2022 [YR2022]`,
               names_to = "year",
               values_to = "amount")|>
  mutate(year = as.integer(year))
copy
```



##Visualize data on a scatter plot
```{r}
library(ggplot2)
scatterplot <- ggplot(data = copy,
       mapping = aes(x = amount,
                     y = `Country Name`)) +
  geom_point() +
  labs(x = "GDP per capita, PPP (current international $)",
       y = "Top 20 GDP countries in the world",
       title = "GDP over time (1990-2022)",
      )

scatterplot


```





##Explanation

I isolates the top 20 GDPs in the world in 2019. Then, I created the data visualization from those values across time.
















