---
title: "Gov 50 Final Project"
author: "Anita Carew"
description: "December 13th, 2023"
output:
  distill::distill_article:
    self_contained: false
---


#Cleaning data set

```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(readr)
library(base)
final_wb_data <- read_csv("final_wb_data.csv")
df <- final_wb_data |>
  janitor::clean_names() |>
  mutate(series_name = if_else(series_name == "CO2 emissions (metric tons per capita)", "CO2", "GDP")) 

#Pivot longer CO2 variable
co2 <- df |>
  filter(series_name == "CO2") |>
  pivot_longer(cols = starts_with("x"),  
               names_to = c("series", "year"),  
               names_pattern = "x(\\d+)_yr(\\d+)",
               values_to = "CO2") |>
  select(country_name, country_code, year, CO2)

#Pivot longer GDP variable
gdp <- df |>
  filter(series_name == "GDP")  |>
  pivot_longer(cols = starts_with("x"),  
               names_to = c("series", "year"),  
               names_pattern = "x(\\d+)_yr(\\d+)",  
               values_to = "GDP") |>
  select(country_name, country_code, year, GDP)
#Left join data
final_df <- left_join(co2, gdp, by = c("country_code", "country_name", "year"))

# Convert specific columns to numeric
final_df$CO2 <- as.numeric(final_df$CO2)
final_df$GDP <- as.numeric(final_df$GDP)

# Get rid of NA values
final_df[final_df == "NA"] <- NA
final_df <- na.omit(final_df)
# or
final_df <- final_df[complete.cases(final_df), ]
final_df$year <- as.numeric(final_df$year)
final_df
```




#Introduction

 My research question for this project is: Does a country's GDP have any effect on its carbon emissions? The widely held belief by political and climate scientists has long been that the higher a country's GDP, the more their carbon emissions increase. This goes along with the idea that if a country is more economically developed, they use more technology to support their quality of lifestyle, and thus their high waste and consumption will lead to higher carbon emissions. There is only a little evidence such as the 2016 report by the International Energy Agency that claims some countries do exist that have been able to reduce Green House Gas (GHG) emissions while growing GDP. Thus, it is important to examine whether this relationship exists, and if so, does a better economy always make our planet suffer?  

My main coefficient of interest is the coefficient associated with the GDP variable in my regression model. My main outcome of interest is CO2 emissions. Therefore, my null hypothesis is that there is no correlation or relationship between GDP and CO2 emissions worldwide. I will reject this hypothesis if I notice any correlation in my visualizations or significance within my regression table.





#Data section

My data is derived from the World Bank. My sample is made up of all the countries the World Bank measures, including some territories (total: 217). The independent variable is GDP per capita, PPP (current international $). This is measured by using the gross domestic product (GDP) expressed in current international dollars converted by purchasing power parity (PPP) conversion factor. The dependent variable is carbon emissions which we will assume rise as a result of higher GDP. This dependent variable is measures by carbon dioxide emissions (metric tons per capita), which is seen through emissions which are stemming from the burning of fossil fuels and the manufacture of cement. They include carbon dioxide produced during consumption of solid, liquid, and gas fuels and gas flaring. 

My research design is cross-sectional. In order to produce a plot that summarizes the main outcome variable, worldwide CO2 emissions, I found the mean of CO2 emissions of every country in the data set for every year and plotted it on a line chart, ranging from 1990-2020. As can be seen below, the plot reflects an increasing trend until around 2007, when emissions began to decline with much variability, a steep decline from 2015 until 2020. 

##First visualization

```{r}
##Aggregate data set
agg_data <- final_df |>
  group_by(year) |>
  summarize(
    CO2 = mean(CO2, na.rm = TRUE),
    GDP = mean(GDP, na.rm = TRUE)
  )

##CO2 plot 
ggplot(data = agg_data,
       mapping = aes(x = year,
                     y = CO2)) +
  geom_point() + geom_smooth() +
labs(x = "Years",
       y = "Average CO2 emissions (metric tons per capita)",
       title = "Worldwide CO2 over time",
       subtitle = "Plot visualizing main outcome of interest",
       source = "World Bank, 1990-2020")

```






#Results section

Below you can see the plot of main analysis, where both CO2 emissions are GDP per capita are measured for every year from 1990-2020. As mentioned before, an aggregate data set was used that found the worldwide mean from each year. Due to the difference of scale between how GDP and CO2 emissions are measured, there are two y-axes. The left side shows average CO2 emissions (metric tons per capita), while the right side represents average GDP per capita, PPP (current international $).

#Second visualization

```{r}
#Graph with only CO2
plot(agg_data$year, agg_data$CO2, type = "l", col = "red", ylab = "CO2", xlab = "Year")
# Adding a second y-axis for GDP on the right side
par(new = TRUE)
plot(agg_data$year, agg_data$GDP, type = "l", col = "blue", yaxt = "n", xaxt = "n", ylab = "", xlab = "")
axis(4)
mtext("GDP", side = 4, line = 3, col = "blue")

legend("topleft", legend = c("CO2", "GDP"), col = c("red", "blue"), lty = 1, cex = 0.8)
title(main = "Worldwide CO2 and GDP over time")

```

##Regression output

```{r}
library(modelsummary)
##Creating new variable for log of GDP
final_df$log_GDP <- log(final_df$GDP)
# Making the regression model
dual_regression <- lm(CO2 ~ GDP + log_GDP, data = final_df)

modelsummary::modelsummary(dual_regression,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```

The plot of main analysis shows that GDP has steadily grown over time. Nevertheless, it does not by itself suggest any correlation between GDP and CO2 emissions. 

When making the regression table, I included the log value of GDP seeing as economic variables like GDP are often right-skewed, as was seen in the visualization. Taking the logarithm compresses the range of values, making the distribution more symmetric and reducing the impact of extreme values. This is useful because we are using worldwide data, and in this case there are a few countries that have significantly higher GDPs than the rest.

The main coefficient of interest is the one associated with log_GDP, which is found to be 0.910 with a standard error of 0.052 and a p-value less than 0.001. This indicates a statistically significant positive relationship between log_GDP and CO2 emissions. In the context of this study, it shows that for every one-unit increase in the log of GDP per capita, there is a 0.910 unit increase in CO2 emissions, holding other variables constant. Overall, all the p=values in this table were p=<0.001, and seeing as a p-value less than 0.05 is often considered statistically significant, it suggests that all coefficients in this table are statistically significant.

The intercept is -6.504, which represents the estimated CO2 emissions when log_GDP and GDP are both zero, though this might not have a practical interpretation in this context because GDP cannot realistically be zero.




#Conclusion section

The statistical significance of the log_GDP coefficient suggests that there is evidence to reject the null hypothesis that the true coefficient is zero. However, we cannot infer causation based on this regression analysis. The positive coefficient implies an association between higher GDP per capita and increased CO2 emissions, but it does not specify the direction of causation. It could be that higher GDP leads to more emissions, but it could also be that countries with higher emissions experience economic growth. Thus, the interpretation should lean towards association rather than causation. In conclusion, the estimated coefficient for log_GDP is statistically significant, indicating a positive association with CO2 emissions, but careful consideration is needed when inferring causality.

One possible limitation is that some data was NA from the data set. In order to complete the calculations and make visualizations, I had to remove these rows, which means there was some missing data. We were also limited by the presence of potential confounding variables not included in the model might affect the observed relationship between GDP and CO2 emissions. For instance, factors like technological advancements, government policies, or changes in energy sources could confound the observed relationship. In addition, the assumption of a linear relationship between GDP and CO2 emissions might oversimplify the underlying dynamics, as the relationship could be nonlinear or subject to threshold effects.

To enhance the analysis, more time and resources could be used to collect improved and comprehensive data. This could involve obtaining detailed information on country-specific policies, technological advancements, and energy efficiency measures over time. Including such variables could help address confounding and provide a more nuanced understanding of the factors influencing CO2 emissions. Furthermore, as touched on in the introduction, GDP might affect CO2 emissions based on the economic level of the country. Thus, this analysis could be improved by dividing the data set to examine the GDP and CO2 relationship in developing countries vs. developed countries.
